Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-15 ms.qmd
Fixing up some comments from Franka dn Shinichi throughout, add new affiliation for Shin, wording changes

Daniel W.A. Noble^1^ $\ddagger$, Fonti Kar^2^, Alex Bush^4^, Frank
Seebacher^3^ $\dagger$, & Shinichi Nakagawa^2^ $\dagger$

@@

^1^ Division of Ecology and Evolution, Research School of Biology, The
Australian National University, Canberra, ACT 2600, Australia\
^2^ Ecology and Evolution Research Centre, School of Biological, Earth
and Environmental Sciences, University of New South Wales, Sydney, NSW,
Australia\
^3^ School of Life and Environmental Sciences, University of Sydney,
Sydney, NSW 2006, Australia\
^4^ Department of Biology, Lancaster University, Liverpool, UK\
$\dagger$ contributed equally\
$\ddagger$ corresponding author, daniel.noble\@anu.edu.au

**Running head**: Reduced physiological rate variation in ectotherms

@@

Climate change is expected to result in warmer and more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is predicted to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Reversible forms of phenotypic plasticity, such as acclimation, are expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted
at higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. NEED MORE ABOUT WHY VARIANCE IS EXPECTED TO CHANGE!

Changes in variability of traits governing energy demand are expected to have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023; @Bolnick2011]. More variable populations, genetically and/or phenotypically, are also  predicted to be associated with broader niches, have increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking;
@forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. In addition, if phenotypic and genetic variation are correlated and traits are linked to fitness, changes in variation may limit responses to selection impacting a populations capacity to evolve [@Pelletier2012]. Therefore, maintaining intrapopulation variability in physiological rates in a warmer world may be important for population resilience to climate change.

@@

@@

  # Contrats adult vs juv across habitats regardless of the acute/acclim
               contrast_lnrr_lh_cat_stats <- rbind(get_lifestage(pred, habitat = "t", newdata),
                                                   get_lifestage(pred, habitat = "m", newdata),
                                                   get_lifestage(pred, habitat = "f", newdata))

@@

### *Simulating among-individual variance in performance curves*
To better understand the characteristics of the performance curves in a sample that would lead to observed changes in variance across temperature we conducted a simple simulation. To simulate performance curves, we used an asymmetrical Gaussian function:

where $T$ is the temperature gradient, $\delta$ is the optimal temperature (the temperature where performance is maximized), $\sigma$ the performance breadth, and $\alpha$ the skewness of the performance function or rate variation (see @fig-sim in *Supplementary Materials* for example curves).  While we could have modelled performance curves differently than @eq-sim (e.g., Arrhenius), our intentions were not to model the exact curves (which are not know) but rather identify what might vary across the major habitat types that could contribute to the different patterns of variance change at higher temperatures we observed across habitats.

We simulated *n* = 1000 individual performance curves by varying the amount of between individual variance on each of the key parameters ($\delta$, $\sigma$, $\alpha$) in all possible combinations from 0.01 to 2. We then took the mean and standard deviation of the performance at two temperatures 10 degrees apart (18 and 28 degrees) to calculate the log variance ratio $lnVR_{Q_{10}}$. Our simulations assume that the value of variance does not change depending on the temperatures choosen, but given our analysis restricts the data to temperatures within the normal thermal range of each species we do not envisage this being problematic.

@@

The final dataset included a total of `r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`,
amphibians = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`,
reptiles = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`,
arthropods = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and nematode species), `r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`, annelids = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, echinoderms = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`, reptiles = `r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and cnidarian species), and `r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (annelids = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
reptiles = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))` and amphibians =
`r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))` along with a single tardigrade species) (@fig-phylo). We had more data on acute thermal responses (n = `r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`) because acute responses were reported for each of the two acclimation temperatures (@fig-phylo). The two acute $lnRR_{Q_{10}}$ effect sizes ([@fig-concept]A) differed significantly from each other ($\beta$ = `r mean(change_acute)`, 95% CI: `r quantile(change_acute, 0.025)` to `r quantile(change_acute, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(change_acute))`) with animals acclimated to high temperatures having slightly higher average $lnRR_{Q_{10}}$ ($\mu$ = `r mean(hot_acute)`, 95% CI: 95% CI: `r quantile(hot_acute, 0.025)` to

@@

```{r, fig-phylo}
#| label: fig-phylo
#| echo: false
#| fig-cap: "Phylogenetic distribution of acute and acclimation $lnRR_{Q_{10}}$/$lnCVR_{Q_{10}}$ estimates across major habitats.  The total number of acute and acclimation  $lnRR_{Q_{10}}$/$lnCVR_{Q_{10}}$ effect sizes are indicated by the coloured bars as well as whether the taxa are marine, freshwater or terrestrial. Silhouettes are only representative taxa of major clades within the tree."

@@

Capacity for acclimation depended on the habitat. Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-2]A) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r abs(mean(change_f))`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r abs(mean(change_m))`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation (possibly even inverse acclimation) -- showing a `r abs(mean(change_t))`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-2]A). All results are from Model 2 (see “Meta-Analysis” above).

@@

Acclimation capacity did not vary consistently by life-history stage with no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles, except for marine habitats where adults appeared to be able to acclimate to a greater extent, although the effect size was small and in the same direction and magnitude of juveniles ("Model 4"; [@fig-cvrlh]A-C). When comparing between acute and acclimation responses between adults and juveniles there were also no differences across habitats (Adult-Juvenile (Acute): `r mean(acute_all_lh)`, 95% CI: `r quantile(acute_all_lh, 0.025)[1]` to `r quantile(acute_all_lh, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acute_all_lh)`; Adult-Juvenile (Acclimation): `r mean(acclim_all_lh)`, 95% CI: `r quantile(acclim_all_lh, 0.025)[1]` to `r quantile(acclim_all_lh, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acclim_all_lh)`).

@@

############
# Full fig
############
figbtrait <-(p3 | p1 | p2) / (p3_vr_trait | p1_vr_trait | p2_vr_trait) + plot_annotation(tag_levels = "A") 
ggsave(figbtrait, filename = "./output/figures/fig5.png", width = 28, height = 14, device = "png")

figbtrait <- magick::image_read("./output/figures/fig5.png")
figbtrait
```

@@

fig4 <-(p3 | p1 | p2) / (p3_vr | p1_vr | p2_vr) + plot_annotation(tag_levels = "A") 
ggsave(fig4, filename = "./output/figures/fig6.pdf", width = 28, height = 14, device = "pdf")

fig4 <- magick::image_read("./output/figures/fig6.pdf")
fig4
```

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B) with less heterogeneity overall ($I^2_{sv}$ = `r mean(tot_i2_VR)`, 95% CI: `r quantile(tot_i2_VR, 0.025)` to `r quantile(tot_i2_VR, 0.975)`; "Model 1"; see *Supplementary Materials* @fig-s3). However, physiological rate variance increased significantly more in freshwater for acute responses (freshwater vs terrestrial: $\beta$ = `r mean(f_t_c_a)`, 95% CI: `r quantile(f_t_c_a, 0.025)` to `r quantile(f_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_a))`), but not for acclimation responses because increases in rates were dampened by acclimation resulting in smaller increases in variance (freshwater vs terrestrial: $\beta$ = `r mean(f_t_c_aclim)`, 95% CI: `r quantile(f_t_c_aclim, 0.025)` to `r quantile(f_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_aclim))`). While marine ectotherms had larger increases in physiological rate variance compared to terrestrial ectotherms these were not significant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater did not differ in how much variance changed at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between $lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ in any habitat ([@fig-2]B). Differences in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary Materials* analyses with $lnCVR_{Q_{10}}$; @tbl-slopes & @fig-mv).

Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms and a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms when temperatures increased by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a `r mean(f_VR_mean_ef)`%, 95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))` increase in variance at high temperatures (@fig-2; results from "Model 2"). The relative differences between habitats were similar when considering $lnCVR_{Q_{10}}$ (See *Supplementary Materials*).

$lnVR_{Q_{10}}$ values from our simulations matched our empirical results quite nicely in key areas of parameter space (@fig-sim2). Interestingly, for a given between individual variance in thermal breadth, terrestrial ectotherms are predicted to have lower levels of between individual variation in thermal optima compared to marine and freshwater ectotherms ([@fig-sim2]). In contrast, terrestrial ectotherms are expected to have higher levels of between individual variation in thermal breadth when controlling for between individual variation in thermal optima ([@fig-sim2])

>>>>

Daniel W.A. Noble^1^ $\ddagger$, Fonti Kar^2^, Alex Bush^4^, Frank
Seebacher^3^ $\dagger$, & Shinichi Nakagawa^2,5^ $\dagger$

@@

^1^ Division of Ecology and Evolution, Research School of Biology, The
Australian National University, Canberra, ACT 2600, Australia\
^2^ Ecology and Evolution Research Centre, School of Biological, Earth
and Environmental Sciences, University of New South Wales, Sydney, NSW,
Australia\
^3^ School of Life and Environmental Sciences, University of Sydney,
Sydney, NSW 2006, Australia\
^4^ Department of Biology, Lancaster University, Liverpool, UK\
^5^ Department of Biological Sciences, University of Alberta, CW 405, Biological Sciences Building, Edmonton, AB T6G 2E9, Canada\
$\dagger$ contributed equally\
$\ddagger$ corresponding author, daniel.noble\@anu.edu.au

**Running head**: Physiological rate variation in ectotherms

@@

Climate change is expected to result in warmer and more variable thermal environments globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Greater thermal variability is predicted to pose strong selection pressure that leads to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one]. Without plasticity or adaptation, high extinction rates are expected unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016].

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to environmental conditions making it the first ‘line-of-defense’ against environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). However, longer-lasting (days-weeks) temperature increases move environmental conditions away from thermal optima of physiological rate functions (i.e., thermal performance curves) can be mitigated by acclimation, which adjust reaction rates [@Seebacher2015-wi; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. Acclimation therefore alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Acclimation is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across habitats (terrestrial, marine and freshwater) [@steele2019comparison], and as such, we expect organisms within these habitats to vary in their capacity for plastic adjustments in physiology. Species occupying terrestrial habitats have been shown to exhibit weak acclimation abilities, and so, are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper; @morley2019physiological]. In contrast, marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, the focus of research up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

As mean physiological rates increase with temperature there is reason to believe that the intrapopulation variability might also be impacted. Positive mean-variance relationships are common across biology suggesting that, as rates increase, so to should variation in rates (i.e., Taylor's Law; []).

Changes in physiological rate variability is expected to have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming; @Sanderson2023; @Bolnick2011]. Generally, more variable populations are predicted to be associated with broader niches, have increased growth rate, and decreased vulnerability to environmental change lowering extinction risk (i.e., "portfolio effects", [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation; @Schindler2010]. In addition, if phenotypic and genetic variation are correlated and traits are linked to fitness, changes in variation may limit responses to selection impacting a populations capacity to evolve [@Pelletier2012]. Therefore, maintaining intrapopulation variability in physiological rates in a warmer world may be important for population resilience to climate change.

@@

### *Identifying patterns of among-individual variance in performance curves contributing to variance increases*
To better understand the characteristics of performance curves among individuals in a sample that would lead to observed changes in variance across temperature we conducted a simple simulation. To simulate performance curves, we used an asymmetrical Gaussian function (@angilletta2009thermal):

where $T$ is the temperature gradient, $\delta$ is the optimal temperature (the temperature where performance is maximized), $\sigma$ the performance breadth, and $\alpha$ the skewness of the performance function or rate variation (see @fig-sim in *Supplementary Materials* for example curves).  We simulated *n* = 1000 individual performance curves by varying the amount of between individual variance on each of the key parameters ($\delta$, $\sigma$, $\alpha$) in all possible combinations from 0.01 to 2. We then took the standard deviation of the performance at two temperatures 10°C apart (18 and 28°C) to calculate $lnVR_{Q_{10}}$ and identify parameter space that matches observed patterns in our empirical data.

@@

  # Contrast adult vs juv across habitats regardless of the acute/acclim
               contrast_lnrr_lh_cat_stats <- rbind(get_lifestage(pred, habitat = "t", newdata),
                                                   get_lifestage(pred, habitat = "m", newdata),
                                                   get_lifestage(pred, habitat = "f", newdata))

@@

@@

The final dataset included a total of `r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, amphibians = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`, reptiles = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`,
and a single crustacean and nematode species), `r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`, annelids = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, echinoderms = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`, reptiles = `r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, and a single crustacean and cnidarian species), and `r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (annelids = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, molluscs = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, arthropods = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, reptiles = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))` and amphibians = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))` along with a single tardigrade species) (@fig-phylo). We had more data on acute thermal responses (n = `r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`) because acute responses were reported for each of the two acclimation temperatures (@fig-phylo). The two acute $lnRR_{Q_{10}}$ effect sizes ([@fig-concept]A) differed significantly from each other ($\beta$ = `r mean(change_acute)`, 95% CI: `r quantile(change_acute, 0.025)` to `r quantile(change_acute, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(change_acute))`) with animals acclimated to high temperatures having slightly higher average $lnRR_{Q_{10}}$ ($\mu$ = `r mean(hot_acute)`, 95% CI: 95% CI: `r quantile(hot_acute, 0.025)` to

@@

```{r, fig-phylo}
#| label: fig-phylo
#| echo: false
#| fig-cap: "Phylogenetic distribution of acute and acclimation $lnRR_{Q_{10}}$/$lnVR_{Q_{10}}$ estimates across major habitats.  The total number of acute and acclimation  $lnRR_{Q_{10}}$/$lnVR_{Q_{10}}$ effect sizes are indicated by the coloured bars as well as whether the taxa are marine, freshwater or terrestrial. Silhouettes are only representative taxa of major clades within the tree."

@@

Capacity for acclimation depended on the habitat. Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-2]A) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r abs(mean(change_f))`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r abs(mean(change_m))`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation with a `r abs(mean(change_t))`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-2]A). All results are from Model 2 (see “Meta-Analysis” above).

@@

Acclimation capacity did not vary consistently by life-history stage with no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles, except for marine habitats where adults appeared to be able to acclimate to a greater extent ("Model 4"; [@fig-cvrlh]A-C). When comparing between acute and acclimation responses between adults and juveniles irrespective of effect type there were also no differences across habitats (Adult-Juvenile: Terrestrial: `r contrast_lnrr_lh_cat_stats[1,3]`, 95% CI: `r contrast_lnrr_lh_cat_stats[1,4]` to `r contrast_lnrr_lh_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[1,6]`; Marine: `r contrast_lnrr_lh_cat_stats[2,3]`, 95% CI: `r contrast_lnrr_lh_cat_stats[2,4]` to `r contrast_lnrr_lh_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[2,6]`; Freshwater: `r contrast_lnrr_lh_cat_stats[3,3]`, 95% CI: `r contrast_lnrr_lh_cat_stats[3,4]` to `r contrast_lnrr_lh_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[3,6]`).

@@

############
# Full fig
############
figbtrait <-(p3 | p1 | p2) / (p3_vr_trait | p1_vr_trait | p2_vr_trait) + plot_annotation(tag_levels = "A") 
ggsave(figbtrait, filename = "./output/figures/fig4.png", width = 28, height = 14, device = "png")

figbtrait <- magick::image_read("./output/figures/fig4.png")
figbtrait
```

@@

fig4 <-(p3 | p1 | p2) / (p3_vr | p1_vr | p2_vr) + plot_annotation(tag_levels = "A") 
ggsave(fig4, filename = "./output/figures/fig5.pdf", width = 28, height = 14, device = "pdf")

fig4 <- magick::image_read("./output/figures/fig5.pdf")
fig4
```

@@

Variance in physiological rates ($lnVR_{Q_{10}}$) showed an increase with increasing temperature across all habitat types ([@fig-2]B) with less heterogeneity overall ($I^2_{sv}$ = `r mean(tot_i2_VR)`, 95% CI: `r quantile(tot_i2_VR, 0.025)` to `r quantile(tot_i2_VR, 0.975)`; "Model 1"; see *Supplementary Materials* @fig-s3). Overall, there was a `r mean(t_VR_mean_ef)`% (95% CI: `r quantile(t_VR_mean_ef, 0.025)[1]` to `r quantile(t_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(t_VR_mean))`) increase in physiological rate variance for terrestrial ectotherms,  a `r mean(m_VR_mean_ef)`% (95% CI:`r quantile(m_VR_mean_ef, 0.025)[1]` to `r quantile(m_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(m_VR_mean))`) increase in variation for marine ectotherms and a `r mean(f_VR_mean_ef)`%, 95% CI: `r quantile(f_VR_mean_ef, 0.025)[1]` to `r quantile(f_VR_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r p_value(pmcmc(f_VR_mean))` increase in variance for freshwater ectotherms across 10°C (@fig-2; results from "Model 2"). The relative differences between habitats were similar when considering $lnCVR_{Q_{10}}$ (See *Supplementary Materials*).

Physiological rate variance increased significantly more in freshwater compared to terrestrial ectotherms for acute responses ($\beta$ = `r mean(f_t_c_a)`, 95% CI: `r quantile(f_t_c_a, 0.025)` to `r quantile(f_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_a))`), but not for acclimation responses because increases in rates were dampened by acclimation resulting in smaller increases in variance ($\beta$ = `r mean(f_t_c_aclim)`, 95% CI: `r quantile(f_t_c_aclim, 0.025)` to `r quantile(f_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_t_c_aclim))`). While marine ectotherms had larger increases in physiological rate variance compared to terrestrial ectotherms these were not significant (Acute: $\beta$ = `r mean(m_t_c_a)`, 95% CI: `r quantile(m_t_c_a, 0.025)` to `r quantile(m_t_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_a))`; Acclimation: $\beta$ = `r mean(m_t_c_aclim)`, 95% CI: `r quantile(m_t_c_aclim, 0.025)` to `r quantile(m_t_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(m_t_c_aclim))`)([@fig-2]B). Marine and freshwater did not differ in how much variance changed at higher temperatures (Acute: $\beta$ = `r mean(f_m_c_a)`, 95% CI: `r quantile(f_m_c_a, 0.025)` to `r quantile(f_m_c_a, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_a))`; Acclimation: $\beta$ = `r mean(f_m_c_aclim)`, 95% CI: `r quantile(f_m_c_aclim, 0.025)` to `r quantile(f_m_c_aclim, 0.975)`, $p_{MCMC}$ = `r p_value(pmcmc(f_m_c_aclim))`). There were no differences between $lnVR_{Q_{10_{acute}}}$ and $lnVR_{Q_{10_{acclimation}}}$ in any habitat ([@fig-2]B). Differences in variance between habitats were driven by different scaling relationships between mean and variance with slopes varying across habitat types (see *Supplementary Materials* analyses with $lnCVR_{Q_{10}}$; @tbl-slopes & @fig-mv).

$lnVR_{Q_{10}}$ values from our simulations matched our empirical results in particular areas of parameter space (@fig-sim2). For a given between individual variance in thermal breadth, terrestrial ectotherms are predicted to have lower levels of between individual variation in thermal optima compared to marine and freshwater ectotherms ([@fig-sim2]). In contrast, terrestrial ectotherms are expected to have higher levels of between individual variation in thermal breadth when controlling for between individual variation in thermal optima ([@fig-sim2])
